---
title: "Potted seedling AMF"
author: "Khanh Ton"
institute: "University of New Hampshire"
date: "Last updated `r format(Sys.time(), '%m/%d/%Y')`"
bibliography: [references.bib,packages.bib]
csl: ecology.csl
documentclass: article
mainfont: Arial
fontsize: 11pt
geometry: margin=1in
linestretch: 1.5
papersize: a4
format: html
output: bookdown::html_document2
toc: true
always_allow_html: true
editor: visual
---

```{r, include=FALSE,message=FALSE,warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)
library(tinytex)
library(shiny)
library(knitcitations)
library(viridis)
library(gt)
library(car)
library(rstatix)
library(kableExtra)
library(trackdown)
trackdown_auth_configure(path = "client.json")
```

## Introduction

#### Fire effects on mycorrhizal fungi colonization

##### Ectomycorrhizal fungi (EcMF)

##### Arbuscular mycorrhizal fungi (AMF)

Besides creating new aboveground conditions for vegetation succession, fire modifies the physical, chemical, and biotic soil environment. This comprises soil heating [@choromanska2002], charcoal and ash deposition [@ball2010], alteration of biogeochemical processes [@chapin2012], and changing the microbial community including arbuscular mycorrhizal dynamics. Previous studies show that whether AMF mycorrhizal colonization increases or decreases post-fire is site- and fire-specific. In a literature review by @cairney2007, fire effects on AMF vary by fire type (prescribed or wildfire), vegetation type, and time elapsed since the fire. Even then, none of these factors is a good predictor of AMF colonization. For example, @bentivenga1991 and @dhillion1988 both focused on prescribed fires in the prairie ecosystem and sampled for soil fungi relatively soon after the burning events, but the former found an increase in root colonization while the latter detected a decrease in spore density and root colonization. Other potential variables at play are burning temperature, soil moisture level, sampling depth, and methods of collecting and analyzing data. Regardless, there is very little research on the interactions between fire and AMF colonization in the temperate forest ecosystem. Fulfilling this knowledge gap can help inform whether forest management using prescribed burns can promote mycorrhizal symbiosis or, if it does not, can outweigh the reduction of mycorrhizal benefits to oak regeneration. It is worth noting that *Q. rubra* is ectomycorrhizal in nature, but seedlings involved in our experiment were grown from acorns in soil mediums containing arbuscular mycorrhizae (see Limitations for further clarifications). AMF colonization occurred in place of EcMF as a result.

### Research questions and hypotheses

This study explored whether fire benefits oak regeneration in terms of promoting AMF colonization of seedling roots in a mesocosm setup. In this experimental design, oak seedlings were grown from acorns in burned and unburned soil treatments for four months before having their roots harvested. Our objective was to compare the extent of AMF root colonization across seedlings grown in these soils. We hypothesized that AMF root colonization in the burned soils would be more extensive than that in the unburned soils as the fungal structures took advantage of the nutrients released by the burn.

## Methods

### Mesocosm experiment setup

Burned soils were collected at the location of the May 2022 Bemis Fire at Crawford Notch State Park (Hart’s Location, NH) in May 2023. The top 15 cm of soil, including both organic (Oa) and mineral (E, Bh, Bs) horizons, was collected from areas within 5 m of a live, co-dominant *Q. rubra*, selecting microsites where the Oi and Oe layer had been thoroughly consumed by the fire. Leaf litter from fall 2022 was excluded from the sample. Control soils were collected from a similar location adjacent to the burned area, following the same procedure after removing the Oi horizon. Each batch of soil was homogenized after collection, removing only large roots and rocks but keeping the majority of fine roots as part of the sample. Half of each batch was heat-treated at 60∘C for 48 hours. They were then mixed 1:1 with PRO-MIX HPCC MYCORRHIZAE, a potting mix containing *Rhizophagus irregularis*. This potting mix was the leftover from another greenhouse experiment from the previous year. 

A total of 955 acorns were collected from two mother trees in Shelburne and Thornton, NH in fall 2022. After a float test to determine viability, 754 were cold-stratified over the winter. In April 2023, acorns were soaked and the 175 individuals that germinated were planted in pots in May. There were 30-40 replicate pots for each of the five soil treatments: (1) burned soil, (2) unburned soil, (3) heat-treated burned soil, (4) heat-treated unburned soil, and (5) potting mix only (control). Seedlings were placed into rows at random under a mesh tent that lets in moderate sunlight. They were rotated once and watered twice weekly.

### Lab processing of collected samples

#### Root scoring for ectomycorrhizal (EcM) fungi colonization

#### Root clearing, staining, and scoring for arbuscular mycorrhizal (AM) fungi colonization

In August 2023, a subsample of 48 seedlings from a total of 175 were harvested for their roots. Root lengths and number of lateral roots were counted for all, and a smaller subset (PVF seedling roots) was also weighed. The root clearing, staining, and scoring protocol was adapted from @cleavitt2008 . Roots were bleached and submerged in 20 ml glass vials containing 10% ethanol solutions. However, the solution was not effective at stopping pathogenic fungal growth because, at the time of analysis, 5 samples were completely unusable considering the vast extent of their fungal colonization. Thirty were in usable condition, in which the solutions remained relatively clear and there was little to no fungal colonization; 13 other were salvageable, their solutions less clear with more or less spread out colonization.

The 30 usable samples were cleared and stained in spring 2024. Root segments were taken out of vials into test tubes, submerged in 10% KOH solutions, and placed in a 94∘C water bath. The solutions were changed every half an hour until they stopped changing color. At this point, the roots became somewhat translucent and light brown in color. After that, they were acidified with 10% HCl solutions for one hour and then stained with 1:1 glycerol-vinegar solutions overnight. The following day, roots were rinsed and kept in 1:1 glycerol-vinegar destaining solutions in 20ml glass vials (Figure 8).

Samples were individually mounted on a gridded microscope slide (grid size = 1 mm\*1 mm) with a cover slip. A compound microscope was used at 200x magnification to count for different AMF structures. It is worth noting that only AMF structures were present in these samples, despite *Q. rubra* being typically ectomycorrhizal. At 100 intersections in each sample taken at 1 mm intervals, one of the following categories was determined: NS - no structure present, A - arbuscules, V - vesicles, C - coils, and MH - mycorrhizal hyphae (e.g. Figure 9).

#### Root EcMF and AMF colonization data

Percent root colonization for each structure (A, V, and C) was calculated as the ratio of the abundance of the specific structure over the total number of observations (O). Percent hyphal colonization was the ratio of the sum of all structures observed (A, V, C, MH) over O. They were then separately analyzed by one-way ANOVA with burning as the main effect. More one-way ANOVA analyses were carried out with percent hyphal colonization as the response variable and individual soil treatment measurements as the factor variables. Additionally, a multiple linear regression was run to tease apart the relationships between percent hyphal colonization and root lateral abundance, root length, root weight, stem length, stem DRC, and stem weight.

## Results

### Soil nutrient contents

The three soil treatments had similar high levels of Mg and very low levels of Pb (Table 1). The control sample was the most basic and contained the most Ca and the least OM, while the unburned sample contained the least amounts of K and P. Percent hyphal colonization did not differ significantly among different levels of nutrients (p\>0.1).

Table 1. Test results of soil treatments: colors signify whether values lie within the optimal range (white - very low, red - low, yellow - optimal, green - high)

|  |  |  |  |  |  |  |  |
|---------|---------|---------|---------|---------|---------|---------|---------|
| **Soil** | **pH** | **Ca (ppm)** | **Mg (ppm)** | **K (ppm)** | **P (ppm)** | **Pb (ppm)** | **%OM** |
| Control | 5.70 | 814.4 | 430 | 202 | 72 | 0.58 | 6.30 |
| Burned | 5.10 | 614.4 | 255 | 184 | 40 | 2.75 | 10.10 |
| Unburned | 5.05 | 441.0 | 235 | 148 | 29 | 2.30 | 10.33 |

### Root mycorrhizal fungal colonization extent

#### EcMF colonization

#### AMF colonization

There were no significant differences in the extent of hyphal colonization among unburned, burned, and control soil treatments (p=0.2). The average hyphal colonization percentages were 13.9±4.3, 15.6±3.9, and 1.0±0.8 respectively (Figure 1). Similarly, arbuscular colonization among soil treatments was not significantly different (p=0.4). The mean arbuscular colonization of roots in unburned soils was 7.4±2.8, in burned soils 6.3±2.1, and in control 0.5±0.3 (Figure 2). Coil colonization also did not significantly differ (p=0.4), with the average extent in unburned soils being 5.2±2.7, burned soils 8.1±2.8, and control 0.5±0.6 (Figure 3).

Percent hyphal colonization was not significantly correlated to the six seedling measurements: root lateral abundance, root length, root weight, stem length, stem DRC, and stem weight (Figures 4 & 5). Hyphal colonization seemed to have a positive relationship with root lateral abundance and stem weight while having an inverse relationship with the remaining variables, but all correlation coefficients were close to 0 (Table 2).

Table 2. Correlation coefficients between AMF hyphal colonization and seedling measurements

|  |  |  |  |  |  |
|------------|------------|------------|------------|------------|------------|
| Root length (mm) | Root weight (g) | Number of laterals | Stem length (mm) | Stem DRC (mm) | Stem weight (g) |
| -0.01 (p=0.95) | -0.27 (p=0.40) | 0.08 (p=0.73) | -0.09 (p=0.67) | -0.08 (p=0.71) | 0.03 (p=0.92) |

Figure 1. AMF hyphal colonization of Q. rubra grown in different soil mixtures (x indicates average colonization extent)

Figure 2. AMF arbuscular colonization of Q. rubra grown in different soil mixtures (x indicates average colonization extent)

Figure 3. AMF coil colonization of Q. rubra grown in different soil mixtures (x indicates average colonization extent)

Figure 4. Correlations between AMF hyphal colonization and Q. rubra root length, weight, and number of laterals across all treatments

Figure 5. Correlations between AMF hyphal colonization and Q. rubra stem length, diameter at root collar, and weight across all treatments

## Discussion

#### Fire effects on mycorrhizal fungal colonization

##### EcMF

##### AMF

Fire can alter soil resource supply by accelerating nutrient cycling and hence releasing more of them [@chapin2012]. It also can consume more volatile nutrients and lead to their decline. Any temperature above 60∘C is found lethal to fine roots and soil organisms [@watts2018]. A high-intensity fire (420 - 1160∘C) can volatilize soil C and OM to a depth of 10 cm while a low-intensity one (40 - 360∘C) can reach 5 cm. The intensity of the Bemis Fire remained unclear besides that it consumed the Oi and Oe layers, but one year seemed to have been a proficient duration for OM and Mg to recover as their abundances were similar between the burned and unburned samples. On the other hand, there were more K and P in the burned sample than the unburned one -- the Bemis Fire could have been responsible for this increase. The low Ca level in both burned and unburned soil samples suggests that perhaps the soils at Crawford Notch State Park are inherently low in Ca. 

The resulting insignificant differences in hyphal colonization extents among different soil treatments rejected our hypothesis that burned soils would have more extensive colonization. A meta-analysis by @dove2017 supported this result, highlighting no significant response in mycorrhizal colonization post-fire across studies, ecosystems, and fire types. However, it is noted that fire effects were dependent on the assessment method used. For example, the in situ approach reported an average of 21% reduction of mycorrhizal colonization while the ex situ bioassays yielded no significance. A reduction in propagule quantity by heating was also observed in an experimental setting by @pattinson1999 as three *Glomus spp.* underwent a simulated fire of moderate intensity. The hyphal network was most severely disturbed at the soil surface and the effect decreased with depth, despite that normally propagules are densest at the top.

In the field, fire effects on mycorrhizal colonization can vary by biomes and fire types. Three studies on shrublands in different parts of the world with similar times elapsed since fire (1-9 months) found different trends: the wildfire in Australia posed no effect on propagules [@bellgard1994], and so did the prescribed burn in south-central Florida (USA) with root colonization [@anderson1997], but the wildfire in Pakistan decreased propagule viability [@rashid1997]. Fire effects may also change temporally, or not -- Alaskan boreal forest sites 3, 15, 45, and 80 years following fire did not see significant differences in AMF abundance [@treseder2004]. Another confounded variable that could have affected the colonization extent is inoculum density. The inoculum density hypothesis stated that a high inoculum density could prevent a decrease in colonization in response to disturbance [@jasper1991]. @mcgonigle2000 disagreed, finding that high inoculum density facilitated the development of a difference in colonization after a disturbance. Our result may have sided with the latter, assuming low inoculum density occurred across all soil treatments given how low the colonization rates were.

Additionally, the average colonization extents of all AMF structures in burned soils and unburned soils were closer to one another than those in the control treatment. Overall, roots in burned soils had the most extensive colonization rate of 16% while the control barely hosted any (1%). When compared by structure, there were more arbuscules in unburned soils than in their burned counterparts, while coils had the opposite trend. This might mean that the *R. irregularis* spores in the potting mix may have played a trivial role in facilitating AMF colonization compared to existing spores in the burned and unburned soils.

Moreover, hyphal colonization was shown to have insignificant relationships with different seedling measurements. When burned and unburned treatments were grouped together, all correlation coefficients hovered around 0. When observed separately, hyphal colonization was positively correlated to root length and stem weight in unburned samples and to the number of laterals and stem length in burned samples, though not significant either. Regardless, the remaining negative relationships carried little meaning and should be deemed negligible. In other words, AMF colonization extent did not seem to be a good indicator of *Q. rubra* seedling growth and health.

### Limitations

#### Mycorrhizal fungi colonization analysis

It is also worth noting that the author of this paper was only involved in the processes of root processing, AMF counting, and data analysis -- the mesocosm experiment setup was carried out by others and the author gave limited input into the experimental design. One major concern before starting the experiment was that the soil collection process may have disturbed the mycelium and thus reduced the fungal population in the samples. Following the experiment, root samples were initially observed for ectomycorrhizal structures, which were absent, so they were then checked for arbuscular structures. As previously mentioned, these seedlings were grown from acorns in potted condition with added AMF spores, which may have favored AMF colonization over EcMF. Perhaps the results could have been further solidified were we able to analyze all the 48 harvested root samples instead of 30 as they are heavily skewed towards the burned and unburned treatments (13 each, and only 4 controls). Better storage techniques are needed to prevent fungal growth on the harvested roots.

## Citations

### R packages

```{r generateBibliography,results="asis",echo=FALSE}
read.bibtex(file="packages.bib")
```

### Literature
