---
title: "Oak at the Edge"
author: "Khanh Ton"
institute: "University of New Hampshire"
date: "Last updated `r format(Sys.time(), '%m/%d/%Y')`"
bibliography: [references.bib,packages.bib]
csl: ecology.csl
documentclass: article
mainfont: Arial
fontsize: 11pt
geometry: margin=1in
linestretch: 1.5
format: html
editor: visual
---

```{r,include=FALSE,message=FALSE,warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)
library(tinytex)
library(shiny)
library(knitcitations)
library(RColorBrewer)
library(viridis)
library(gt)
library(car)
library(multcomp)
library(rstatix)
library(kableExtra)
```

## Results

### Stand characterization

#### Understory competition

In burned stands, *Q. rubra* mainly competes with early successional seedlings e.g. *Rubus spp.* `RUS`, *A. rubrum* `ACRU`, *B. alleghaniensis* `BEAL` and stump sprouts e.g. *F. grandifolia* `FAGR`.

```{r,include=FALSE,message=FALSE,warning=FALSE}
comp <- read_csv("competition.csv")

#Add Treatment & Pair columns
comp2 <- comp %>% mutate(Treatment=case_when(
  Unit %in% c("3/16","5/15","28/2","12/2") ~ "Shelterwood",
  Unit %in% c("20/2","3/1") ~ "Seedtree",
  Unit %in% c("44B","45C","46B","46C") ~ "Clearcut"))

comp2 <- comp2 %>% replace_na(list(
  Unit="Unnamed",Treatment="Unmanaged"))

comp2 <- comp2 %>% mutate(Pair=case_when(
  Unit %in% c("3/16","5/15")~"1",
  Unit %in% c("20/2","20/3","20/4","20/5","3/1")~"2",
  Unit %in% c("28/2","28/3","12/2")~"3",
  Unit=="Unnamed"~"4",
  Unit %in% c("44B","45C")~"5",
  Unit %in% c("46B","46C")~"6"))

comp2$Spcode <- replace(comp2$Spcode,
                        comp2$Spcode %in% c("RUID","RUFR","RUOC"),
                        "RUS")

comp2 <- comp2 %>% filter(Spcode %in% 
                            c("ABBA","ACPE","ACRU","ACSA",
                              "BEAL","BEPA","BEPO","FAGR",
                              "FRAM","OSVI","PIST","POGR",
                              "POTR","PRPE","RUS","SACA",
                              "TSCA"))

comp2 <- comp2 %>% 
  mutate(Location=as.factor(Location),
         Unit=as.factor(Unit),
         Disturbance=as.factor(Disturbance),
         Treatment=as.factor(Treatment),
         Pair=as.factor(Pair),
         Spcode=as.factor(Spcode))

#spcode count
count <- comp2 %>% group_by(Pair,Disturbance,Spcode) %>% 
  summarize(stemcount=sum(Count))

#scale up to per ha
b1 <- 66
c1 <- 46
b2 <- 31
c2 <- 30
b3 <- 30
c3 <- 45
b4 <- 43
c4 <- 19
b5 <- 10 
c5 <- 13
b6 <- 28
c6 <- 32

plotarea <- 0.0001

count <- count %>% mutate(Density=if_else(
  Pair=="1" & Disturbance=="B",
  stemcount/b1/plotarea,if_else(Pair=="1" & Disturbance=="C",
  stemcount/c1/plotarea,if_else(Pair=="2" & Disturbance=="B",
  stemcount/b2/plotarea,if_else(Pair=="2" & Disturbance=="C",
  stemcount/c2/plotarea,if_else(Pair=="3" & Disturbance=="B",
  stemcount/b3/plotarea,if_else(Pair=="3" & Disturbance=="C",
  stemcount/c3/plotarea,if_else(Pair=="4" & Disturbance=="B",
  stemcount/b4/plotarea,if_else(Pair=="4" & Disturbance=="C",
  stemcount/c4/plotarea,if_else(Pair=="5" & Disturbance=="B",
  stemcount/b5/plotarea,if_else(Pair=="5" & Disturbance=="C",
  stemcount/c5/plotarea,if_else(Pair=="6" & Disturbance=="B",
  stemcount/b6/plotarea,stemcount/c6/plotarea)))))))))))) 

count$Density <- round(count$Density,0)

count <- count %>% group_by(Pair, Disturbance) %>% 
  slice_max(Density,n=3)
```

```{r,echo=FALSE,fig.cap="Stem density (per ha) of understory species in study stands"}
#visualize stem density
bc <- c("B"="#CC6677","C"="#88CCEE")
p6 <- c("1"="#FBE3D6","2"="#FFFFCC","3"="#E8E8E8","4"="#C2F1C8","5"="#DCEAF7","6"="#DCEAF7")

count %>% 
ggplot(aes(x=Spcode,y=Density))+
  geom_rect(aes(fill=Pair),alpha=0.5,
            xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf)+
  geom_col(aes(fill=Disturbance),width=0.6,
           position=position_dodge(width=0.7))+
  scale_fill_manual(values=c(p6,bc))+
  facet_wrap(vars(Pair),scales='free')+
  theme_light()+
  theme(text=element_text(size=8.5))+
  xlab("Species")+
  ylab("Stem count (per ha)")
```

#### Overstory composition

\[commentary here\]

```{r, include=FALSE,message=FALSE,warning=FALSE}
trees <- read_csv("overstory.csv")

#Add Treatment & Pair columns
trees2 <- trees %>% mutate(Treatment=case_when(
  Unit %in% c("3/16","5/15","28/2","12/2") ~ "Shelterwood",
  Unit %in% c("20/2","3/1") ~ "Seedtree",
  Unit %in% c("44B","45C","46B","46C") ~ "Clearcut"))

trees2 <- trees2 %>% replace_na(list(
  Unit="Unnamed",Treatment="Unmanaged"))

trees2 <- trees2 %>% mutate(Pair=case_when(
  Unit %in% c("3/16","5/15")~"1",
  Unit %in% c("20/2","20/3","20/4","3/1")~"2",
  Unit %in% c("28/2","12/2","12/3","12/4","12/5","12/6",
              "12/7","12/8","12/9","12/10","12/11")~"3",
  Unit=="Unnamed"~"4",
  Unit %in% c("44B","45C")~"5",
  Unit %in% c("46B","46C")~"6"))

trees2 <- trees2 %>% mutate(Location=as.factor(Location)) %>% 
  mutate(Location=fct_recode(Location,
                             "BEF"="Bartlett Experimental Forest",
                             "CF"="Crawford Notch State Park",
                             "HOG"="Hogsback (Blueberry Mountain)",
                             "SB"="Stevens Brook (Plummers Ledge)"))

trees2 <- trees2 %>% 
  mutate(Location=as.factor(Location),
         Unit=as.factor(Unit),
         Disturbance=as.factor(Disturbance),
         Treatment=as.factor(Treatment),
         Pair=as.factor(Pair),
         Spcode=as.factor(Spcode))

#dbh classes
trees2 <- trees2 %>% mutate(dbhClass=if_else(
  `DBH (cm)`<=5,"2-5",if_else(`DBH (cm)`>5 & `DBH (cm)`<=10,
                "5-10",if_else(`DBH (cm)`>10 & `DBH (cm)`<=20,
                "10-20",if_else(`DBH (cm)`>20 & `DBH (cm)`<=30,
                "20-30",if_else(`DBH (cm)`>30 & `DBH (cm)`<=40,
                "30-40",if_else(`DBH (cm)`>40 & `DBH (cm)`<=50,
                "40-50",if_else(`DBH (cm)`>50 & `DBH (cm)`<=60,
                "50-60",">60"))))))))

trees2 <- trees2 %>% mutate(dbhClass=fct_relevel(dbhClass,"2-5","5-10","10-20",
                                                "20-30","30-40","40-50","50-60",">60"))

#basal areas
trees2 <- trees2 %>% mutate(ba=0.00007854*(`DBH (cm)`^2))

sumba <- trees2 %>% select(Pair,Disturbance,Spcode,dbhClass,ba)
sumba <- sumba %>% group_by(Pair,Disturbance,Spcode,dbhClass) %>% 
  summarize(sumba=sum(ba))

#scale up to per ha
b1 <- 66
c1 <- 46
b2 <- 31
c2 <- 30
b3 <- 30
c3 <- 45
b4 <- 43
c4 <- 19
b5 <- 10 
c5 <- 13
b6 <- 28
c6 <- 32

plotarea <- 0.00785398163397448

sumba <- sumba %>% mutate(m2ha=if_else(
  Pair=="1" & Disturbance=="B",
  sumba/b1/plotarea,if_else(Pair=="1" & Disturbance=="C",
  sumba/c1/plotarea,if_else(Pair=="2" & Disturbance=="B",
  sumba/b2/plotarea,if_else(Pair=="2" & Disturbance=="C",
  sumba/c2/plotarea,if_else(Pair=="3" & Disturbance=="B",
  sumba/b3/plotarea,if_else(Pair=="3" & Disturbance=="C",
  sumba/c3/plotarea,if_else(Pair=="4" & Disturbance=="B",
  sumba/b4/plotarea,if_else(Pair=="4" & Disturbance=="C",
  sumba/c4/plotarea,if_else(Pair=="5" & Disturbance=="B",
  sumba/b5/plotarea,if_else(Pair=="5" & Disturbance=="C",
  sumba/c5/plotarea,if_else(Pair=="6" & Disturbance=="B",
  sumba/b6/plotarea,sumba/c6/plotarea)))))))))))) 

sumba <- sumba %>% filter(Spcode %in% c("ABBA","ACPE","ACRU","ACSA",
                                        "BEAL","BEPA","BEPO","FAGR",
                                        "FRAM","OSVI","PIRU","PIST",
                                        "POGR","POTR","PRPE","PRSE",
                                        "QURU","TIAM","TSCA"))
```

```{r,echo=FALSE, fig.cap="Basal area (per ha) of overstory species in study stands by DBH size classes"}
#let's try to plot this lol
sp <- c( "ABBA"="#000004FF","ACPE"= "#08071EFF","ACRU"= "#180F3EFF",
         "ACSA"= "#2D1160FF","BEAL"= "#451077FF","BEPA"= "#5C167FFF",
         "BEPO"="#721F81FF","FAGR"= "#882781FF","FRAM"= "#9F2F7FFF",
         "OSVI"="#B63679FF","PIRU"= "#CD4071FF","PIST"= "#E24D66FF",
         "POGR"= "#F1605DFF","POTR"= "#F9795DFF","PRPE"= "#FD9567FF",
         "PRSE"= "#FEAF77FF","QURU"= "#FEC98DFF","TIAM"= "#FDE3A5FF","TSCA"= "#FCFDBFFF")

sumba %>% 
  ggplot(aes(x=dbhClass,y=m2ha))+
  geom_rect(aes(fill=Pair),alpha=0.5,
            xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf)+
  geom_col(aes(fill=Spcode))+
  scale_fill_manual(values=c(p6,sp))+
  facet_grid(rows=vars(Pair),cols=vars(Disturbance),scales='free')+
  theme_light()+
  theme(text=element_text(size=8.5))+
  xlab("DBH size classes (cm)")+
  ylab("Basal area (per ha)")

#we have not taken into account dead/alive trees
```

### Oak seedling density and measurements

#### 2023 oak seedling density

Seedling density increased threefold in burned stands (2359+/-211 per ha) relative to control stands (778+/-121 per ha, p\<0.001).

```{r,include=FALSE,message=FALSE,warning=FALSE}
count23 <- read_csv("oakcount5m2023.csv")

#Add Treatment & Pair columns
count23x <- count23 %>% mutate(Treatment=case_when(
  Unit %in% c("3/16","5/15","28/2","12/2") ~ "Shelterwood",
  Unit %in% c("20/2","3/1") ~ "Seedtree",
  Unit %in% c("44B","45C","46B","46C") ~ "Clearcut"))

count23x <- count23x %>% replace_na(list(
  Unit="Unnamed",Treatment="Unmanaged"))

count23x <- count23x %>% mutate(Pair=case_when(
  Unit %in% c("3/16","5/15")~"1",
  Unit %in% c("20/2","20/3","20/4","20/5","3/1")~"2",
  Unit %in% c("28/2","28/3","12/2")~"3",
  Unit=="Unnamed"~"4",
  Unit %in% c("44B","45C")~"5",
  Unit %in% c("46B","46C")~"6"))

count23x <- count23x %>% mutate(Location=as.factor(Location)) %>% 
  mutate(Location=fct_recode(Location,
                             "BEF"="Bartlett Experimental Forest",
                             "CF"="Crawford Notch State Park",
                             "HOG"="Hogsback (Blueberry Mountain)",
                             "SB"="Stevens Brook (Plummers Ledge)"))

count23x <- count23x %>% 
  mutate(Unit=as.factor(Unit),
         Disturbance=as.factor(Disturbance),
         Treatment=as.factor(Treatment),
         Pair=as.factor(Pair))

plotarea <- 0.00785398163397448

#summarize data
count23x <- count23x %>% na.omit() %>% 
  mutate(Density=OAKcount/plotarea)

count23x$Density <- round(count23x$Density,0)
```

```{r,echo=FALSE,message=FALSE}
count23b <- count23x %>% filter(Disturbance=="B")
stats_count23b <- get_summary_stats(count23b)
stats_count23b <- stats_count23b %>% mutate(Disturbance="B")
stats_count23b <- stats_count23b %>% filter(variable=="Density") %>% select(Disturbance,min,max,median,mean,sd,se)

count23c <- count23x %>% filter(Disturbance=="C")
stats_count23c <- get_summary_stats(count23c)
stats_count23c <- stats_count23c %>% mutate(Disturbance="C")
stats_count23c <- stats_count23c %>% filter(variable=="Density") %>% select(Disturbance,min,max,median,mean,sd,se)

stats_count23 <- stats_count23b %>% full_join(stats_count23c)
stats_count23 %>% kable(caption="Summarized statistics of 2023 oak seedling density per ha") %>% kable_styling(html_font="Arial",bootstrap_options = c("striped", "hover","condensed", "responsive"))
```

```{r,echo=FALSE,warning=FALSE,fig.cap="2023 oak seedling density per ha of study stands"}
#visualize stem density
count23x %>% filter(Pair==c("1","2","3","4")) %>% 
  ggplot(aes(x=Pair,y=Density,fill=Disturbance))+
  geom_rect(aes(fill=Pair),alpha=0.5,
            xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf)+
  geom_boxplot(width=0.6,position=position_dodge(width=0.7))+
  scale_fill_manual(values=c(p6,bc))+
  facet_grid(~Pair,scales='free_x')+
  theme_minimal()+
  ylab("Stem count (per ha)")+
  labs(fill="Pairs")
```

```{r,include=FALSE}
#levenes test
leveneTest(count23x$Density~count23x$Disturbance)

#anova
count23x <- count23x %>% filter(Pair %in% c("1","2","3","4"))
one.way <- aov(OAKcount~Disturbance+Pair,data=count23x)
summary(one.way)
```

#### 2023 diameter at root collar (DRC, mm)

DRC was greater for seedlings in burned stands (4.6+/-0.3 mm) versus control stands (3.3+/-0.3 mm, p\<0.01).

```{r,include=FALSE,message=FALSE,warning=FALSE}
ms23 <- read_csv("oakmsms23.csv")

ms23 <- ms23 %>% mutate(Location=as.factor(Location)) %>% 
  mutate(Location=fct_recode(Location,
                             "CF"="Crawford Notch State Park",
                             "HOG"="Hogsback (Blueberry Mountain)",
                             "SB"="Stevens Brook (Plummers Ledge)"))

#Add Treatment & Pair columns
ms23x <- ms23 %>% mutate(Treatment=case_when(
  Unit %in% c("3/16","5/15","28/2","12/2") ~ "Shelterwood",
  Unit %in% c("20/2","3/1") ~ "Seedtree",
  Unit %in% c("44B","45C","46B","46C") ~ "Clearcut"))

ms23x <- ms23x %>% replace_na(list(
  Unit="Unnamed",Treatment="Unmanaged"))

ms23x <- ms23x %>% mutate(Pair=case_when(
  Unit %in% c("3/16","5/15")~"1",
  Unit %in% c("20/2","3/1")~"2",
  Unit %in% c("28/2","12/2")~"3",
  Unit=="Unnamed"~"4"))

ms23x <- ms23x %>% 
  mutate(Unit=as.factor(Unit),
         Disturbance=as.factor(Disturbance),
         Treatment=as.factor(Treatment),
         Pair=as.factor(Pair))
```

```{r,echo=FALSE,message=FALSE}
ms23b <- ms23x %>% filter(Disturbance=="B")
stats_ms23b <- get_summary_stats(ms23b)
stats_ms23b <- stats_ms23b %>% mutate(Disturbance="B")
stats_ms23b <- stats_ms23b %>% 
  select(Disturbance,variable,min,max,median,mean,sd,se) %>% 
  filter(variable %in% c("Height_cm","DRC_mm",
                         "nlive_branches","ndead_branches"))

ms23c <- ms23x %>% filter(Disturbance=="C")
stats_ms23c <- get_summary_stats(ms23c)
stats_ms23c <- stats_ms23c %>% mutate(Disturbance="C")
stats_ms23c <- stats_ms23c %>% 
  select(Disturbance,variable,min,max,median,mean,sd,se) %>% 
  filter(variable %in% c("Height_cm","DRC_mm",
                         "nlive_branches","ndead_branches"))

stats_ms23 <- stats_ms23b %>% full_join(stats_ms23c)
stats_ms23 %>% 
  kable(caption="Summarized statistics of 2023 oak seedling measurements") %>%
  kable_styling(html_font="Arial",bootstrap_options = c("striped", "hover","condensed", "responsive"))
```

```{r,echo=FALSE,warning=FALSE,fig.cap="2023 seedling DRCs by study stands"}
ms23x %>% 
  ggplot(aes(x=Pair,y=DRC_mm,fill=Disturbance))+
  geom_rect(aes(fill=Pair),alpha=0.5,
            xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf)+
  geom_boxplot(width=0.6,position=position_dodge(width=0.7))+
  scale_fill_manual(values=c(p6,bc))+
  facet_grid(~Pair,scales='free_x')+
  theme_minimal()+
  ylab("2023 diameter at root collar (mm)")+
  labs(fill="Pairs")
```

```{r,include=FALSE}
leveneTest(ms23x$DRC_mm~ms23x$Disturbance)

aov_drc23 <- aov(DRC_mm~Disturbance+Pair,data=ms23x)
summary(aov_drc23)
```

#### 2024 extension growth (cm)

Extension growth was greater for seedlings in burned stands (6.43+/-0.5 cm) versus control stand (2.6+/-0.4 cm, p\<0.001).

```{r,include=FALSE}
ms24 <- read_csv("oakmsms24.csv")

ms24 <- ms24 %>% mutate(Location=as.factor(Location)) %>% 
  mutate(Location=fct_recode(Location,
                             "CF"="Crawford Notch State Park",
                             "HOG"="Hogsback (Blueberry Mountain)",
                             "SB"="Stevens Brook (Plummers Ledge)"))

ms24x <- ms24 %>% mutate(Treatment=case_when(
  Unit %in% c("3/16","5/15","28/2","12/2") ~ "Shelterwood",
  Unit %in% c("20/2","3/1") ~ "Seedtree",
  Unit %in% c("44B","45C","46B","46C") ~ "Clearcut"))

ms24x <- ms24x %>% replace_na(list(
  Unit="Unnamed",Treatment="Unmanaged"))

ms24x <- ms24x %>% mutate(Pair=case_when(
  Unit %in% c("3/16","5/15")~"1",
  Unit %in% c("20/2","3/1")~"2",
  Unit %in% c("28/2","12/2")~"3",
  Unit=="Unnamed"~"4"))

ms24x <- ms24x %>% 
  mutate(Unit=as.factor(Unit),
         Disturbance=as.factor(Disturbance),
         Treatment=as.factor(Treatment),
         Pair=as.factor(Pair),
         Survey=as.factor(Survey))

ms24x <- ms24x %>% drop_na()

ms24x <- ms24x %>% mutate(herbivory_pct=na_if(herbivory_pct,"na")) %>% 
  mutate(Pathogen_damage_pct=na_if(Pathogen_damage_pct,"na"))
```

```{r, echo=FALSE,message=FALSE}
ms24b <- ms24x %>% filter(Disturbance=="B")
stats_ms24b <- get_summary_stats(ms24b)
stats_ms24b <- stats_ms24b %>% mutate(Disturbance="B")
stats_ms24b <- stats_ms24b %>% 
  select(Disturbance,variable,min,max,median,mean,sd,se) %>% 
  filter(variable %in% c("Height_cm","Extension_growth_cm","DRC_mm",
                         "nleaves","nlive_branches","ndead_branches"))

ms24c <- ms24x %>% filter(Disturbance=="C")
stats_ms24c <- get_summary_stats(ms24c)
stats_ms24c <- stats_ms24c %>% mutate(Disturbance="C")
stats_ms24c <- stats_ms24c %>% 
  select(Disturbance,variable,min,max,median,mean,sd,se) %>% 
  filter(variable %in% c("Height_cm","Extension_growth_cm","DRC_mm",
                         "nleaves","nlive_branches","ndead_branches"))

stats_ms24 <- stats_ms24b %>% full_join(stats_ms24c)
stats_ms24 %>% kable(caption="Summarized statistics of 2024 oak seedling measurements") %>% kable_styling(html_font="Arial",bootstrap_options = c("striped", "hover","condensed", "responsive"))
```

```{r,echo=FALSE,fig.cap="2024 seedling extension growths by study stands"}
ms24x %>%
  ggplot(aes(x=Pair,y=Extension_growth_cm,fill=Disturbance))+
  geom_rect(aes(fill=Pair),alpha=0.5,
            xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf)+
  geom_boxplot(width=0.6,position=position_dodge(width=0.7))+
  scale_fill_manual(values=c(p6,bc))+
  facet_grid(~Pair,scales='free_x')+
  theme_minimal()+
  ylab("2024 extension growth (cm)")+
  labs(fill="Pairs")
```

```{r,include=FALSE}
leveneTest(ms24x$Extension_growth_cm~ms24x$Disturbance)

aov_exgr24 <- aov(Extension_growth_cm~Disturbance+Pair+Survey,data=ms24x)
summary(aov_exgr24)
```

#### 2024 number of live branches

There were more live branches per seedling in the burned stands than the control stands, respectively 4+/-0 and 2+/-0 (p\<0.001).

```{r,echo=FALSE,fig.cap="2024 number of live branches per seedling by study stands and surveys"}
ms24x %>% filter(Survey=="1") %>% 
  ggplot(aes(x=Pair,y=nlive_branches,fill=Disturbance))+
  geom_rect(aes(fill=Pair),alpha=0.5,
            xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf)+
  geom_boxplot(width=0.6,position=position_dodge(width=0.7))+
  scale_fill_manual(values=c(p6,bc))+
  facet_grid(~Pair,scales='free_x')+
  theme_minimal()+
  ylab("2024 number of live branches, survey 1")+
  labs(fill="Pairs")

ms24x %>% filter(Survey=="2") %>% 
  ggplot(aes(x=Pair,y=nlive_branches,fill=Disturbance))+
  geom_rect(aes(fill=Pair),alpha=0.5,
            xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf)+
  geom_boxplot(width=0.6,position=position_dodge(width=0.7))+
  scale_fill_manual(values=c(p6,bc))+
  facet_grid(~Pair,scales='free_x')+
  theme_minimal()+
  ylab("2024 number of live branches, survey 2")+
  labs(fill="Pairs")
```

```{r,include=FALSE}
leveneTest(ms24x$nlive_branches~ms24x$Disturbance)

aov_lb24 <- aov(nlive_branches~Disturbance+Pair+Survey,data=ms24x)
summary(aov_lb24)
```

#### 2024 number of leaves

Seedlings in burned stands sprouted more leaves (16+/-1) than in control stands (7+/-1, p\<0.001).

```{r,echo=FALSE,fig.cap="2024 number of leaves per seedling by study stands"}
ms24x$nleaves <- replace(ms24x$nleaves,ms24x$nleaves=="100+","100") %>%
  as.numeric()

ms24x %>% 
  ggplot(aes(x=Pair,y=nleaves,fill=Disturbance))+
  geom_rect(aes(fill=Pair),alpha=0.5,
            xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf)+
  geom_boxplot(width=0.6,position=position_dodge(width=0.7))+
  scale_fill_manual(values=c(p6,bc))+
  facet_grid(~Pair,scales='free_x')+
  theme_minimal()+
  ylab("2024 number of leaves")+
  labs(fill="Pairs")
```

```{r,include=FALSE}
leveneTest(ms24x$nleaves~ms24x$Disturbance)

aov_leaves24 <- aov(nleaves~Disturbance+Pair+Survey,data=ms24x)
summary(aov_leaves24)

tukey <- TukeyHSD(aov_leaves24)
tukey

#Extracting significant pairs
tukey_df <- as.data.frame(tukey$Pair)
names(tukey_df)[names(tukey_df) == "p adj"] <- "p.adj"
sig_pairs <- subset(tukey_df,p.adj < 0.05)
tukey_df

# Create list of comparisons from significant pairs
comparisons <- strsplit(row.names(sig_pairs), "-")
comparisons
```

### Leaf Area Index (LAI)

Burned stands have lower LAI values (averaging 3.2+/-0.2) than control stands (5.4+/-0.2, p\<0.001).

```{r,include=FALSE,message=FALSE,warning=FALSE}
lai <- read_csv("LAIR.csv")

#Add Treatment & Pair columns
lai2 <- lai %>% mutate(Treatment=case_when(
  Unit %in% c("3/16","5/15","28/2","12/2") ~ "Shelterwood",
  Unit %in% c("20/2","3/1") ~ "Seedtree",
  Unit %in% c("44B","45C","46B","46C") ~ "Clearcut"))

lai2 <- lai2 %>% replace_na(list(
  Unit="Unnamed",Treatment="Unmanaged"))

lai2 <- lai2 %>% mutate(Pair=case_when(
  Unit %in% c("3/16","5/15")~"1",
  Unit %in% c("20/2","3/1")~"2",
  Unit %in% c("28/2","12/2")~"3",
  Unit=="Unnamed"~"4",
  Unit %in% c("44B","45C")~"5",
  Unit %in% c("46B","46C")~"6"))

lai2 <- lai2 %>% 
  mutate(Location=as.factor(Location),
         Unit=as.factor(Unit),
         Disturbance=as.factor(Disturbance),
         Treatment=as.factor(Treatment),
         Pair=as.factor(Pair))
```

```{r,echo=FALSE,message=FALSE}
laib <- lai2 %>% filter(Disturbance=="B")
stats_laib <- get_summary_stats(laib)
stats_laib <- stats_laib %>% mutate(Disturbance="B")
stats_laib <- stats_laib %>% filter(variable=="avgLAI") %>% 
  select(Disturbance,min,max,median,mean,sd,se)

laic <- lai2 %>% filter(Disturbance=="C")
stats_laic <- get_summary_stats(laic)
stats_laic <- stats_laic %>% mutate(Disturbance="C")
stats_laic <- stats_laic %>% filter(variable=="avgLAI") %>% 
  select(Disturbance,min,max,median,mean,sd,se)

stats_lai <- stats_laib %>% full_join(stats_laic)
stats_lai %>% kable(caption="Summarized statistics of LAI values") %>% kable_styling(html_font="Arial",bootstrap_options = c("striped", "hover","condensed", "responsive"))
```

```{r,echo=FALSE,fig.cap="LAI values by study stands"}
#LAI! plot by pair and burn y/n
bc <- c("B"="#CC6677","C"="#88CCEE")
p6 <- c("1"="#FBE3D6","2"="#FFFFCC","3"="#E8E8E8",
        "4"="#C2F1C8","5"="#DCEAF7","6"="#DCEAF7")

lai2 %>% 
  ggplot(aes(Pair,avgLAI))+
  geom_rect(aes(fill=Pair),alpha=0.5,
            xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf)+
  geom_boxplot(aes(fill=Disturbance))+
  scale_fill_manual(values=c(p6,bc))+
  facet_grid(~Pair,scales='free_x')+
  theme_minimal()+
  ylab("Average LAI values")
```

```{r,include=FALSE}
leveneTest(lai2$avgLAI~lai2$Disturbance)

aov_lai <- aov(avgLAI~Disturbance+Pair,data=lai2)
summary(aov_lai)

tukey <- TukeyHSD(aov_lai)
tukey

#Extracting significant pairs
tukey_df <- as.data.frame(tukey$Pair)
names(tukey_df)[names(tukey_df) == "p adj"] <- "p.adj"
sig_pairs <- subset(tukey_df,p.adj < 0.05)
tukey_df

# Create list of comparisons from significant pairs
comparisons <- strsplit(row.names(sig_pairs), "-")
comparisons
```

## Citations

### R packages

```{r generateBibliography,results="asis",echo=FALSE}
read.bibtex(file="packages.bib")
```
