---
title: "Oak at the Edge"
author: "Khanh Ton"
bibliography: "references.bib"
format: html
editor: visual
---

## Introduction

### *Quercus rubra* regeneration landscape at present

*Quercus rubra* L. (hereafter *Q. rubra*) is an economically and ecologically important tree species in the northeastern United States. It is a fast-growing species and thrives on a wide range of upland soils \[\@abramsFireDevelopmentOak1992\]. Additionally, it is tolerant of heat, drought, and even ground fire \[\@abramsFireEcologicalHistory2000\]. *Q. rubra* produces high-quality timber with high wood density, making it an important lumber commodity and prompting forest managers to be concerned with its sustainable regeneration \[\@deyReviewDynamicsPrescribed2018\]. *Q. rubra* is managed for other ecological purposes as well. For example, it serves as a habitat and food source for many wildlife species, from birds like turkeys and jays to mammals like mice, squirrels, and deer, and even insects like weevils \[\@mcsheaForestryMattersDecline2007\]. *Q. rubra* drives their population dynamics with its high mast production occurring about every two years.

*Q. rubra* currently faces many regeneration challenges in the northeastern United States, and its recruitment will continue failing without timely and appropriate interventions. One such challenge is acorns becoming a major hard mast source for wildlife consumption since the demise of American chestnut and the spread of beech bark disease. This high seed predation deters successful advance regeneration outside of mast years \[\@mcsheaForestryMattersDecline2007\]. Moreover, if seeds are buried under a thick litter layer, the seed radicle may struggle to reach the mineral soil, and this can negatively affect epicotyl emergence \[\@arthurRefiningOakFireHypothesis2012\]. Even when seedlings are established, they risk getting eaten, especially by deer \[\@mcsheaForestryMattersDecline2007\]. *Q. rubra* is also susceptible to some fungal pathogens including *Phytophthora* and *Armillaria*, which cause sudden oak death and Armillaria root rot respectively. Lastly, its shade intolerance retards its seedling recruitment when there is insufficient light in the understory \[\@nowackiDemiseFireMesophication2008\]. Seedling growth might be slow due to competition from more shade-tolerant species such as *Acer rubrum* and *Fagus grandifolia*. 

### Fire-oak hypothesis

\[\@abramsFireDevelopmentOak1992\] hypothesized that prescribed fire might be crucial for oak regeneration. Fire has been integral to upland oak systems in the eastern United States for millennia. Pre-settlement fires occurred as a function of lightning strikes as well as native activities such as cooking, heating, seedbed preparation, hunting, ceramic manufacture, and communication. Burning continued with European settlement and pitched oak as the dominant species in periodically burned areas \[\@abramsFireDevelopmentOak1992\]. Over time, *Q. rubra* became much more adapted to a periodic fire regime than other hardwoods and, as a result, can take advantage of the postfire environment. However, fire suppression became mainstream policy in the 1920s and promoted forest mesophication \[\@nowackiDemiseFireMesophication2008\]. As the eastern landscape becomes wetter and more shady, mesic microenvironmental conditions improve and favor the regeneration of shade-tolerant species. This positive feedback loop shrinks suitable habitats for shade-intolerant, fire-adapted species like *Q. rubra* over time. Therefore, it has been argued that it is necessary to bring back fire in a controlled manner to reverse the feedback loop and restore the upland oak ecosystems \[\@abramsFireDevelopmentOak1992\].

## Methods

### Site Characteristics

Field measurements were conducted at four locations across the White Mountains National Forest (WMNF) in northern New Hampshire: Crawford Notch State Park (CF), Mount Hogsback (HOG), Stevens Brook (SB), and Bartlett Experimental Forest (BEF).

CF (44.18 North, 71.39 West)

-   mean annual precipitation: 1476 mm (<https://ncei-normals-mapper.rcc-acis.org/>)

-   soil classification:

-   forest composition

HOG (44.04 North, 71.94 West)

-   mean annual precipitation: 1074 mm

-   soil classification

-   forest composition

SB (43.84 North, 71.88 West)

-   mean annual precipitation: 1029 mm

-   soil classification

-   forest composition

BEF (44.06 North, 71.30 West)

-   mean annual precipitation: 1405 mm

-   soil classification

-   forest composition

### Experimental Design

### Samples and Measurements

### Statistical Analysis

## Results

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

### Leaf Area Index (LAI)

```{r}
lai <- read_csv("LAIR.csv")

#Add Treatment & Pair columns
lai2 <- lai %>% mutate(Treatment=case_when(
  Unit %in% c("3/16","5/15","28/2","12/2") ~ "Shelterwood",
  Unit %in% c("20/2","3/1") ~ "Seedtree",
  Unit %in% c("44B","45C","46B","46C") ~ "Clearcut"))

lai2 <- lai2 %>% replace_na(list(
  Unit="Unnamed",Treatment="Unmanaged"))

lai2 <- lai2 %>% mutate(Pair=case_when(
  Unit %in% c("3/16","5/15")~"1",
  Unit %in% c("20/2","3/1")~"2",
  Unit %in% c("28/2","12/2")~"3",
  Unit=="Unnamed"~"4",
  Unit %in% c("44B","45C")~"5",
  Unit %in% c("46B","46C")~"6"))

lai2 <- lai2 %>% 
  mutate(Location=as.factor(Location),
         Unit=as.factor(Unit),
         Disturbance=as.factor(Disturbance),
         Treatment=as.factor(Treatment),
         Pair=as.factor(Pair))
```

```{r}
#LAI! plot by pair and burn y/n
bc <- c("B"="#CC6677","C"="#88CCEE")
p3 <- c("1"="#FBE3D6","2"="#FFFFCC","3"="#E8E8E8")

lai2 %>% filter(Pair %in% c("1","2","3")) %>% 
  ggplot(aes(Pair,avgLAI))+
  geom_rect(aes(fill=Pair),alpha=0.5,
            xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf)+
  geom_boxplot(aes(fill=Disturbance))+
  scale_fill_manual(values=c(p3,bc))+
  facet_grid(~Pair,scales='free_x')+
  theme_minimal()+
  #theme(text=element_text(size=30))+
  ylab("Average LAI values")
```

## Discussion

## Citations
